---
title: How many children under 5 worldwide?
author: Yang Liu
date: '2024-12-13'
slug: how-many-children-under-5-worldwide
categories:
  - Data Visualization
output:
  blogdown::html_page:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE,
                      error = FALSE, fig.width = 12, fig.height = 6)
# error=TRUE allows error
```

```{r}
library("data.table")
library("ggplot2")
library("cowplot")

CURRENT_YEAR <- 2024 # also label on the plot value for the current year

if(dir.exists(here::here("../Data"))){
dt_wpp_world <- fread(here::here("../Data/WPP2024_World_PopulationByAge5GroupSex_Medium.csv"))
source(here::here("Code/prepare_WPP_data.R"))
} else {
dt_wpp_world <- fread("../Data/WPP2024_World_PopulationByAge5GroupSex_Medium.csv")
source("Code/prepare_WPP_data.R")
}
# melt into long-format by sex
dt_wpp_world_l <- melt(dt_wpp_world, id.vars = c("Location", "Time", "AgeGrp", "AgeGrpStart"), 
                       variable.name = "Sex", value.name = "Pop")
dt_wpp_world_l[, Sex := gsub("Pop", "", Sex)]


pw  <- plot.wpp.data(data_plot = dt_wpp_world_l, "Global population")
p5  <- plot.wpp.data(data_plot = dt_wpp_world_l[AgeGrpStart %in% c(0)], title0 = "Global Population of Children Under Age 5 ")
p10 <- plot.wpp.data(data_plot = dt_wpp_world_l[AgeGrpStart %in% c(0:10)], "Global population of children under 15 years old")
p15 <- plot.wpp.data(data_plot = dt_wpp_world_l[AgeGrpStart %in% c(0:15)], "Global population under 20 years old")
p20 <- plot.wpp.data(data_plot = dt_wpp_world_l[AgeGrpStart %in% c(0:20)], "Global population under 25 years old")
p30 <- plot.wpp.data(data_plot = dt_wpp_world_l[AgeGrpStart %in% c(0:30)], "Global population under 35 years old")
p40 <- plot.wpp.data(data_plot = dt_wpp_world_l[AgeGrpStart %in% c(0:40)], "Global population under 45 years old")
p55 <- plot.wpp.data(data_plot = dt_wpp_world_l[AgeGrpStart %in% c(0:55)], "Global population under 60 years old")
p65 <- plot.wpp.data(data_plot = dt_wpp_world_l[AgeGrpStart %in% c(0:65)], "Global population under 70 years old")
p75 <- plot.wpp.data(data_plot = dt_wpp_world_l[AgeGrpStart %in% c(0:75)], "Global population under 80 years old")

gg_world1 <- plot_grid(p5, p10, p15, p20, ncol = 2)
gg_world2 <- plot_grid(p30, p40, p55, p65, p75, pw, ncol = 2)


p5_s  <- plot.wpp.data(data_plot = dt_wpp_world_l[AgeGrpStart %in% c(0)], 
                       title0 = "Global Population of Children Under Age 5 ", by_sex = TRUE)
p10_s <- plot.wpp.data(data_plot = dt_wpp_world_l[AgeGrpStart %in% c(0:10)],
                       title0 = "Global population of children under 15 years old", by_sex = TRUE)
p15_s <- plot.wpp.data(data_plot = dt_wpp_world_l[AgeGrpStart %in% c(0:15)],
                       title0 = "Global population under 20 years old", by_sex = TRUE)
p20_s <- plot.wpp.data(data_plot = dt_wpp_world_l[AgeGrpStart %in% c(0:20)],
                       title0 = "Global population under 25 years old", by_sex = TRUE)
gg_world1_by_sex <- plot_grid(p5_s, p10_s, p15_s, p20_s, ncol = 2)

```

Since these figures are not directly available via Google search, I calculated them using the mid-year population estimates from the World Population Prospects (WPP) 2024: 

**World Population:** In 2024, the global population is estimated at 8.2 billion (8,162 million). The population is projected to peak in 2084 at 10.3 billion.

**Children Under 5 Years Old:** The number of children under 5 is 647 million in 2024. This figure peaked in 2017 at 698 million and is projected to decline to 550 million by 2100.

**Children Aged 0–14 Years:** The population of children aged 0–14 peaked in 2020 and stands at 2.0 billion in 2024.

**Children and Adolescents Aged 0–19 Years:** The population of those aged 0–19 years is 2.7 billion (2,668 million) in 2024, which marks the peak year for this age group.

**Population Under 25 Years Old (0–24):** The number of people aged 0–24 years is 3.3 billion in 2024. This population is expected to experience a slight increase and peak in 2030.

```{r, fig.width = 10, fig.height = 8}
gg_world1

```

**Also by sex:  **

```{r, fig.width = 10, fig.height = 10}
gg_world1_by_sex

```


**World population**  

```{r, fig.width = 10, fig.height = 12}
gg_world2

```

Source: [World Population Prospects (WPP) 2024](https://population.un.org/wpp/Download/Standard/CSV/). Population on 01 July, by 5-year age groups.
