---
title: Notes on writing R package
author: Yang Liu
date: '2019-07-28'
slug: some-experience-on-writing-r-package
categories:
  - R package
tags: []
output:
  blogdown::html_page:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

Although [**'SHAPforxgboost'**](https://github.com/liuyanguu/SHAPforxgboost) is not a complicated package, it took me some time to get the package pass all the cran check. 

I started with using 'usethis' package [https://usethis.r-lib.org/](https://usethis.r-lib.org/) to set up the structure of the package. 

## Description

Single quote packages in both Title and Description fields.For example, this package uses the SHAP values output from 'xgboost'.

## Namespace: 

* It is unnecessary to add **Depends** in the **DESCRIPTION** except `R (>= 3.3.0)`. And the **Imports** part in **DESCRIPTION** is irrelevant to what you import into NAMESPACE in the code (although they are both named _import_).  The NAMESPACE is defined by using `@import` to import packages and functions in the R code. 

  * If you have `@import ggplot2` anywhere in the R code. Your function can use ggplot as long as your package is loaded. Your function knows to search in the namespace of `ggplot2::`, but the ggplot2 package is not **attached**. If someone loads your package, he would needs `library(ggplot2)` if to use ggplot2 in his code. This is the difference between **attach** and **load** as discussed in Hadley's R package book. 

  * By the way, if you put `ggplot2` in **Depends** in the **DESCRIPTION**, `library(yourPackage)` would really attach `ggplot2`. But basically, you don't have to do that for others, since your function can run as long you have included `ggplot2` in the namespace (which you have to do anyway).

* If you name a function like `plot.shap.summary` it would be documented as S3 method even if you `@export plot.shap.summary` (Yes, you have to export the full function name, otherwise, it would be an S3 method in the namespace.). Don't name function this way to avoid confusion with S3method. For example, I changed the name to `shap.plot.summary`.

* Each dataset that you attached using `use_data` need a document as well. Document it in a way similar to the function.

* All the function parameters should be documented using `@param`.  

* When **devtools** check the program, there will be notes that "no visible binding for global variable". This is how to please the cran:

```{r, eval = F}
if(getRversion() >= "2.15.1")  {
  utils::globalVariables(c(".", "rfvalue", "value", ... )) # all the variables complained
  }
```


## Files locations:

* All the codes go into the **R/** folder  

* `@example R/example/sample1.R` will attach **sample1.R** in the folder "R/example/" to the documentation of the function. If you write them directly in the function, you use @example**s** instead of @example.  

